%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% FILE    : pthread-Tutorial.tex
% AUTHOR  : (C) Copyright 2012 by Peter C. Chapin
% SUBJECT : Tutorial on POSIX threads.
%
% This document attempts to give a quick overview of how to use the facilities of POSIX threads.
% It is not intended to be a complete description of pthreads. Instead I only hope to introduce
% the reader to the basic pthread operations. The hope is that there will be enough information
% here to allow the reader to start using pthreads effectively without overwhelming the reader
% with details.
%
% TODO:
%
% + Add some kind of reference or bibliography section.
%
% + Add some kind of glossary and/or list of abbreviations.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%+++++++++++++++++++++++++++++++++
% Preamble and global declarations
%+++++++++++++++++++++++++++++++++
\documentclass{article}
\usepackage{listings}
%\usepackage[dvips]{graphics}
\usepackage{hyperref}
\setlength{\parindent}{0em}
\setlength{\parskip}{1.75ex plus0.5ex minus0.5ex}

% I want to typeset newly introduced terms in a special way.
\newcommand{\newterm}[1]{\emph{#1}}

% Used for bits of C in-line.
\newcommand{\snippet}[1]{\texttt{#1}}

%++++++++++++++++++++
% The document itself
%++++++++++++++++++++
\begin{document}

% The following are settings for the listings package.
\lstset{language=C,
        basicstyle=\small,
        stringstyle=\ttfamily,
        commentstyle=\ttfamily,
        xleftmargin=0.25in,
        showstringspaces=false}

%-----------------------
% Title page information
%-----------------------
\title{pthread Tutorial}
\author{\copyright\ Copyright 2012 by Peter C. Chapin}
\date{September 5, 2012}
\maketitle

\tableofcontents

\section*{Legal}
\label{sec:legal}

\textit{Permission is granted to copy, distribute and/or modify this document under the terms of
  the GNU Free Documentation License, Version 1.1 or any later version published by the Free
  Software Foundation; with no Invariant Sections, with no Front-Cover Texts, and with no
  Back-Cover Texts. A copy of the license is included in the file \texttt{GFDL.txt} distributed
  with the \LaTeX\ source of this document.}

\section{Introduction}
\label{sec:intro}

This document is intended to be a short but useful tutorial on how to use POSIX threads
(pthreads). In this document I do not attempt to give a full description of all pthread
features. Instead I hope to give you enough information to use pthreads in a basic, yet
effective way. Please refer to a text on pthreads for the more esoteric details of the standard.

In addition to talking about the pthread interface itself, I also spend time in this document
discussing issues regarding concurrent programming in general. While such issues are not
specific to pthreads, it is a must that you understand them if you are to use pthreads---or any
thread library---effectively.

I will assume that you are compiling pthreads programs on a Unix system. However, you should be
aware that the pthreads interface is not necessarily specific to Unix. It is a standard
application program interface that could potentially be implemented on many different systems.
However, pthreads is the usual way multi-threaded support is offered in the Unix world. Although
many systems support their own internal method of handling threads, virtually every Unix system
that supports threads at all offers the pthreads interface.

The pthreads API can be implemented either in the kernel of the operating system or in a
library. It can either be preemptive or it can be non-preemptive. A portable program based on
pthreads should not make any assumptions about these matters.

When you compile a program that uses pthreads, you will probably have to set special options on
the compiler's command line to indicate extra (or different) libraries and/or alternative code
generating stratagies. Consult your compiler's documentation for more information on this. Often
you can indicate your desire to use pthreads by supplying the ``-pthread'' option at the end of
the compiler command line. For example

\begin{verbatim}
$ gcc -o myprog myprog.c -pthread
\end{verbatim}

This single option specifies that the pthreads library should be linked and also causes the
compiler to properly handle the multiple threads in the code that it generates.

\input{create-destroy.tex}
\input{sync.tex}
\input{models.tex}
\input{safety.tex}
\input{rules.tex}

\end{document}
